Type theory in type theory with single substitutions

### 2025-01-26 cpp/p.tex
stuff

In this paper, we will answer this question by defining type theory as
a generalised algebraic theory (GAT
\cite{DBLP:journals/apal/Cartmell86}). This has the following
properties.
\begin{enumerate}
\item[(i)] It is an intrinsic presentation
  \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
  opposed to extrinsic
  \cite{alti:phd93,abel2013normalization,DBLP:conf/cpp/AdjedjLMPP24}. That
  is, we only consider well-formed, well-scoped and well-typed
  abstract syntax trees, there are no meaningless terms.
\item[(ii)] The equations of the algebraic theory describe the
  (symmetric) conversion relation.
\end{enumerate}
For type theory, such an intrinsic quotiented presentation is very
natural as typing depends on conversion. The algebraic approach avoids
lots of boilerplate coming from using an explicit conversion relation
(as in e.g.\ \cite{chapman09eatitself}). In the algebraic approach,
syntax and semantics are concordant: the syntax is simply the initial
model, which always exists for any GAT
\cite{DBLP:journals/pacmpl/KaposiKA19}. Such a syntax is quotiented
which means that any function defined on the syntax needs to respect
conversion. For example, it is not possible to define a printing
function which distinguishes the terms $(\lambda x.x)\,1$ and $1$, as
these are convertible. However, it is still possible to define
normalisation \cite{lmcs:4005} and typechecking \cite{typecheck} at
this level of abstraction.

The algebraic/intrinsic approach is also more abstract than extrinsic
approaches, which means that we don't have to make ad-hoc choices. For
example, we do not have to decide whether we want function space {\`a}
la Curry or Church, whether we work with paranoid or economic rules
\cite[Section 5]{winterhalter:hal-01626651}, or whether we have
explicit or recursively defined substitution (instantiation of
variables by terms). These choices are forced by the notion of GAT:
the domain of a lambda appears in its type, se we are always {\`a} la
Church; algebraic theories don't allow recursively defined operators,
so instantiation is an explicit operation, and so on. However, there
are still some options for the exact definition of the substitution
calculus: most notions of type theory use parallel substitutions,
which means that the instantiation operation replaces all variables in
a context by terms in another context \emph{at the same time}. Some
examples of algebraic definitions of type theory using parallel
substitutions ranging from the more syntactic to the more semantic:
Ehrhard's notion \cite{ehrhard,coquandEhrhard}, categories with
families (CwF \cite{DBLP:conf/types/Dybjer95,Castellan2021}), Awodey's
natural models \cite{DBLP:journals/mscs/Awodey18}, contextual
categories \cite{DBLP:journals/apal/Cartmell86}, C-systems
\cite{DBLP:journals/lmcs/AhrensLV18}, locally cartesian closed
categories \cite{DBLP:journals/mscs/ClairambaultD14}, path categories
\cite{10.1145/3204492}. In traditional extrinsic presentations
intantiation replaces one variable by a single term. To bridge the gap
between the intrinsic parallel substitution world and the extrinsic
single substitution world, Voevodsky introduced B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} which are an algebraic
desription of a single substitution calculus. B-systems involve
complex rules describing telescopes, weakenings and substitutions
under telescopes, with several equations.

In this paper, we introduce a new single substitution calculus which
is simpler and more minimalistic than the previous algebraic
approaches. It shows that to explain type theory in an algebraic way,
there is no need for categories, parallel substitutions or weakenings,
empty substitution, telescopes, or combinations of these. All our
operations are easy to motivate, we showcase this in Section
\ref{sec:tt} which is a tutorial introduction to our theory. Our
single substitution calculus is in a sweet spot: our notations read
like traditional extrinsic syntactic presentations, but they are
algebraic, thus come with a well behaved model theory, and can be
easily related to more semantic descriptions.

Our single substitution theory is in some sense too minimalistic: for
example, the equation $b[\p^+][\langle\q\rangle] = b$ is not
\emph{derivable} in any model, but it is \emph{admissible} in the
syntax. (Here $b : \Tm\,(\Gamma\ext A)\,B$ is a term which depends on
a context $\Gamma$ and an extra variable of type $A$; $b[\p^+] :
\Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p^+])$ is a version where we
weakened \emph{before} the last variable; in
$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ we substituted
the last variable for the previous one.)  This is analogous to
e.g.\ parametricity results \cite{DBLP:journals/jfp/BernardyJP12},
which do not hold in an arbitrary model, but they hold in the
syntax. In this paper, we show that the \emph{syntax} of our single
substitution theory and the syntax of the corresponding parallel
substitution theory (CwF-based theory) are isomorphic. All CwF-based
models are models of our single substitution theory, but not
necessarily the other way. While B-systems are equivalent to CwFs (and
C-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more
single substitution models than CwF-based models. This relationship is
similar to that of extensional combinatory calculus and lambda
calculus, where the syntaxes are equivalent, and every lambda model is
a combinatory model, but not the other way round
\cite{DBLP:conf/fscd/AltenkirchKSV23}. A simpler example is the
relationship between monoids over a set $X$ and nil/cons algebras over
$X$, where every monoid is a nil/cons-algebra, but not the other way;
but it is well-known that free monoids over $X$ (syntax of monoid over
$X$) are the same as lists of $X$ (the syntax of nil/cons algebras
over $X$).

Technology to show that syntaxes are equivalent: alpha normalisation
-- terms which don't include instantiation, but are still quotiented.

Minimisation: (i) if we have a universe representing every type as a
term, we can remove some equation, obtaining a more minimalistic, but
isomorphic GAT. (ii) if we even have Sigma tyes, we can actually
derive a parallel model from an SSC model.

SOGAT stuff

The techniques that we use for this proof are not specific to the
SOGAT (\ref{eq:tt}), we expect that they work for any
SOGAT. Isomorphism of the syntaxes means that the corresponding sorts
($\Con$, $\Ty$, $\Tm$) are isomorphic. For a general SOGAT, there are
more models of the corresponding single substitution GAT than the
parallel substitution GAT.

xxx

For formalisation on a computer, the algebraic approach also has its
downsides: as mutually defined recursive operations are not allowed,
substitution (instantiation of variables by terms) is explicit, which
means that it is a constructor, and does not compute by
definition. For example, $(t\,a)[x\mapsto u] = (t[x\mapsto
  u])\,(a[x\mapsto u])$ is a derivable equation in the syntax, but it
is not definitional. This makes formalisation difficult, and results
in the situation called ``transport hell''.

Related work:
- SOGAT->GAT conversion
- B systems
- others

Type theory is a language which can be described as a second-order
generalised algebraic theory (SOGAT
\cite{DBLP:conf/fscd/KaposiX24}). Some properties of this description:

For example, type theory with $\Pi$ types and universes {\`a} la
Coquand \cite{coquandUniverse} is specified by the following SOGAT.
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
\end{alignedat}
\end{equation}
Because types are isomorphic to certain terms (witnessed by $\El$ and $\cd$),
we have to index types by their (universe) level, and types at lower
levels are included at upper levels by $\Lift$. We use implicit
arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly, and $\app$ takes
$i$, $A$ and $B$ implicitly. $f : X \cong Y : g$ denotes an
isomorphism, that is, $f : X\ra Y$, $g:Y\ra X$ with $g\,(f\,x) = x$
and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$ and $\lam$ are
second-order functions. We say that $\Pi$ binds a $\Tm$-variable
in its second $\Ty$-argument. Every binder is represented by
a second-order function, and as a result, we don't need to talk about
contexts and substitutions when describing the theory: we re-use the
binding features of the metatheory to describe those of our object
theory (this technique is also known as higher-order abstract syntax
\cite{DBLP:conf/lics/Hofmann99}, logical framework
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2} or
two-level type theory \cite{DBLP:conf/csl/AltenkirchCK16,DBLP:journals/mscs/AnnenkovCKS23}). There is no good notion of homomorphism between
second-order models, thus the semantics of a SOGAT is given by a first
order generalised algebraic theory (GAT
\cite{DBLP:journals/apal/Cartmell86}). GATs are generalisations of
single-sorted algebraic theories (such as monoids, groups) to multiple
sorts where later sorts can be indexed over previous sorts. The usual
results from universal algebra (e.g.\ models form a category with
finite limits and initial object, existence of free and co-free
models) carry over to GATs \cite{andras,DBLP:phd/hal/Moeneclaey22}. An
example of a GAT is the theory of categories where there is a sort of
objects and a sort of morphisms double-indexed over objects.

Kaposi and Xie \cite{DBLP:conf/fscd/KaposiX24} present two different
first order semantics for any SOGAT: one based on parallel
substitutions, and one based on single substitutions.

In the parallel semantics, the GAT always starts with a category with
a terminal object, objects of the category are called contexts,
morphisms are substitutions. Every sort/operation of the SOGAT gives
rise to a sort/operation in the GAT which is now indexed by
contexts. The sorts moreover come with instantiation (sometimes
called substitution) operations which are functorial, and all the
operations are natural with respect to instantiation. For those sorts
which are bound by an operation in the SOGAT description, we have
context extension, meaning that contexts can carry variables of that
sort (in other words, the sort is given by a locally representable
presheaf). The parallel semantics of the SOGAT
\begin{equation}\label{eq:tytm}
\begin{alignedat}{10}
  & \Ty && : \Set \\
  & \Tm && : \Ty \ra\Set 
\end{alignedat}
\end{equation}
is the GAT of categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}, Definition \ref{def:cwf}).
We define the notion ``model of the SOGAT'' by its parallel substitution
semantics, which is a GAT, and that has an obvious notion of model. The
syntax of the language is the initial model, which exists for every
GAT \cite{DBLP:journals/pacmpl/KaposiKA19}. This is an intrinsic
syntax (only well-typed terms), quotiented by the equations in the
SOGAT, which describe conversion for the language.

The single substitution semantics is a GAT which does not start with a
category: there are contexts and maps between contexts, but these maps
are single substitutions or single weakenings, they don't
compose. There are fewer operations and fewer equations than in the
parallel semantics. In this paper we investigate the single
substitution semantics of the type theory specified by the SOGAT
(\ref{eq:tt}). This presentation of type theory is simpler than CwFs,
the operations are all easy to motivate, we illustrate this in Section
\ref{sec:tt}. Single substitution calculi are closer to the usual
presentation of type theory using derivation rules, this was one of
the motivations of Voevodsky for developing B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. Our single substitution
semantics of (\ref{eq:tytm}) is simpler and has more models than
B-systems.

When showing that the parallel and single substitution syntaxes are
equivalent, we define the parallel operations by induction on the
single substitution syntax. One of the operations in the single
substitution syntax is the instantiation operation: $\blank[\blank] :
\Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta$, and similarly for
terms. When doing induction on types/terms, we have to provide
a method for these operations, but this is difficult as
substitutions don't compose. This is why we derive a new induction
principle which works on $\alpha$-normal types and terms. Just as
$\beta$-normal forms don't distinguish between $\beta$-equal terms,
$\alpha$-normal forms don't include explicit substitutions.
$\alpha$-normal forms are not usual normal
forms in the sense that they are still quotiented by the
computation/uniqueness rules like $\beta$/$\eta$ for function
space. To derive the new induction principle for the syntax, we first
prove $\alpha$-normalisation.

In the particular theory (\ref{eq:tt}), every type is represented by a
term. In its single substitution calculus, we use this to adapt some
equations and remove some others, obtaining a more minimalistic, but
isomorphic GAT.

Theory (\ref{eq:tt}) can be extended
with unit and $\Sigma$ types, more precisely with the following rules.
\begin{equation}\label{eq:sigma}
  \begin{alignedat}{10}
    & \top && : \Ty\,0 \\
    & \tt && : \Tm\,\top \\
    & \top\eta && : t = \tt \\
    & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
    & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
\end{alignedat}
\end{equation}
Using the same insight about types and the universe, the fact that
dependencies can be represented by function spaces, and that
collections of dependencies can be collected using $\Sigma$ types, we
can actually derive a parallel model from any single substitution
model. This way we show that for the particular theory (\ref{eq:tt})+(\ref{eq:sigma}),
single and parallel substitution models are actually equivalent in a
weak sense.

To summarise, our contributions are the following:
\begin{itemize}
\item A new generalised algebraic presentation of type theory in the
  form of a minimalistic single substitution calculus. We show that
  there is no need for parallel substitutions, empty substitution,
  parallel weakenings, telescopes, or combinations of these to
  describe the syntax of type theory in an intrinsic way. We present
  our calculus in an easy-to-understand way.
\item The $\alpha$-normalisation technique and deriving parallel
  substitutions from the single substitution syntax.
\item For type theory with $\Pi$, $\U$, a minimised presentation of
  the equations which results in an isomorphic theory.
\item For type theory with $\Pi$, $\U$, $\top$, $\Sigma$, we show that
  single and parallel models are equivalent in a weak sense.
\end{itemize}

@@

In Section \ref{sec:tt}, we introduce the single substitution calculus
(SSC) only relying on a working knowledge of an implementation of type
theory such as Agda, no prior experience with the metatheory of type
theory is required. Section \ref{sec:admissible} shows that in the
syntax of SSC, several new equations are admissible, and that the
syntax of SSC is actually isomorphic to the CwF-based syntax. In
Section \ref{sec:minimisation} we minimise our SSC obtaining a theory
with fewer equations, relying on the presence of $\Pi$ and $\U$ in our
theory. In Section \ref{sec:cwf}, we show that in the presence of
$\Pi$, $\U$ and $\Sigma$, SSC-models are actually equivalent to
CwF-models in a weak sense. We conclude in Section
\ref{sec:conclusion}.

@@

In the 1990s single substitution calculi for type theory were popular,
but usually in the extrinsic setting, see e.g.\ the thesis of
Altenkirch \cite{alti:phd93}.

B-systems were introduced by Voevodsky \cite{voevodsky2014bsystems} as

@@

substitutions. B-systems correspond to our single substitution
calculus as essentially algebraic theories correpond to generalised
algebraic theories, or sets with a map into $I$ correspond to indexed
families over $I$ \cite[page
  221]{DBLP:journals/apal/Cartmell86}. Another difference is that we
have fewer and less general equations, resulting in the fact that we
have more models than B-systems. However in the syntax of our theory,
all the rules of B-systems are admissible. We describe the

@@

The first generalised algebraic presentation of type theory was Thomas
Ehrhard's calculus \cite{ehrhard,coquandEhrhard} which featured a
parallel substitution calculus with $\p$, $\blank^+$ and
$\langle\blank\rangle$ operations, just like in our calculus. Our
single substitution calculus is essentially Ehrhard's calculus with
the categorical composition and identity operations
removed. Categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}) feature $\p$, $\q$,
$\blank,\blank$ operations which make it more apparent that
substitutions are lists of terms. CwFs are equivalent to Ehrhard's
calculus, and so are contextual categories/C-systems
\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
the natural models of Awodey \cite{DBLP:journals/mscs/Awodey18} and
B-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.

Most computer formalisations of type theory are extrinsic
\cite{DBLP:journals/pacmpl/0001OV18,DBLP:conf/cpp/AdjedjLMPP24,DBLP:journals/jar/SozeauABCFKMTW20}:
they define the syntax as abstract syntax trees, and equip it with
typing and conversion relations. A higher level representation is
working in setoid hell \cite{chapman09eatitself}: terms are
intrinsically well-typed, but conversion is still an explicit
relation. For formalising the GAT-level syntax, one needs a stronger
metatheory than ordinary Agda or Coq: the metatheory has to support
quotient inductive-inductive types (QIITs
\cite{DBLP:journals/pacmpl/KaposiKA19}), in other words, initial
models of GATs. Altenkirch and Kaposi
\cite{DBLP:conf/popl/AltenkirchK16} formalised the syntax of type
theory using postulated QIITs in Agda, together with parametricity and
normalisation \cite{lmcs:4005} proofs. Brunerie and de Boer
\cite{initiality-agda} constructed the initial contextual category in
Agda using postulated quotients. QIITs are supported by the cubical
set model \cite{DBLP:conf/lics/CoquandHM18} and the setoid model
\cite{kaposi-qiit-setoid}, thus Cubical Agda
\cite{DBLP:journals/jfp/VezzosiMA21} and setoid/observational type
theory \cite{setoid,DBLP:phd/hal/Pujet22} support QIITs. For example,
in Cubical Agda, set-truncated and groupoid-truncated syntaxes of type
theory have been formalised \cite{cohtt}. Without QIITs, tricks such
as shallow embedding the syntax can be used to formalise metatheoretic
results about type theory such as gluing
\cite{kaposi_et_al:LIPIcs:2019:10532}, and its special cases
canonicity and parametricity \cite{kaposi-shallow}.

At the level of abstraction of SOGATs
\cite{uemura,DBLP:conf/fscd/KaposiX24}, the difference between single
and parallel substitution calculi are not visible. However, the
metatheory of type theory can be still investigated in this abstract
setting, via the methods of Synthethic Tait Computability
\cite{DBLP:phd/us/Sterling22} or internal sconing
\cite{DBLP:conf/fscd/BocquetKS23}.

% TODO: say that this is the paper version of the \cite{singleTypes}
% paper

>>>>

In this paper, we will answer this question by defining type theory as
a generalised algebraic theory (GAT
\cite{DBLP:journals/apal/Cartmell86}). GATs are multi-sorted algebraic
theories where later sorts can be indexed by previous sorts. An
example is categories where the sort of morphisms is double-indexed
over objects.

The GAT presentation of type theory is intrinsic
\cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} rather
than extrinsic
\cite{alti:phd93,abel2013normalization,DBLP:conf/cpp/AdjedjLMPP24}:
this means that we only consider well-formed, well-scoped and
well-typed abstract syntax trees, there are no meaningless terms; in
the GAT presentation, the conversion relation is expressed by
equations of the algebraic theory.

Some advantages of using a GAT presentation:
\begin{itemize}
\item  In type theory, typing depends on conversion:
  in an extrinsic presentation, this is expressed by a rule such as
  the following.
  \[
  \infer{\Gamma\vdash t : B}{A \sim B && \Gamma\vdash t : A}
  \]
  When conversion is equality, this rule holds by the general
  properties of equality; and we also avoid lots of boilerplate
  stating that e.g.\ conversion is a congruence with respect to all
  operations (for an intrinsic presentation with explicit conversion
  relation see \cite{chapman09eatitself}).
\item Syntax and semantics are concordant: the syntax is simply the
  initial model, which always exists for any GAT and is called a
  quotient inductive-inductive type
  \cite{DBLP:journals/pacmpl/KaposiKA19}. The fact that the syntax is
  quotiented means that any function defined on it needs to respect
  the equations. For example, it is not possible to define a printing
  function which distinguishes the terms $(\lambda x.x)\,1$ and $1$,
  as these are convertible. However, it is still possible to define
  normalisation \cite{lmcs:4005} and typechecking \cite{typecheck} at
  this level of abstraction.
\item The GAT approach is also more abstract than extrinsic
  approaches, which means that we don't have to make ad-hoc
  choices. For example, we do not have to decide whether we want
  function space {\`a} la Curry or {\`a} la Church, whether we work
  with paranoid or economic typing rules \cite[Section
    5]{winterhalter:hal-01626651}, or whether we have explicit or
  recursively defined substitution (instantiation of variables by
  terms, written $t[x\mapsto u]$ or $t[u/x]$).  These choices are
  forced by the notion of GAT: the domain of a lambda appears in its
  type, se we are always {\`a} la Church; algebraic theories don't
  allow recursively defined operators, so instantiation is an explicit
  operation, and so on.
\end{itemize}

% For formalisation on a computer, the algebraic approach also has its
% downsides: as mutually defined recursive operations are not allowed,
% substitution (instantiation of variables by terms) is explicit, which
% means that it is a constructor, and does not compute by
% definition. For example, $(t\,a)[x\mapsto u] = (t[x\mapsto
%   u])\,(a[x\mapsto u])$ is a derivable equation in the syntax, but it
% is not definitional. This makes formalisation difficult, and results
% in the situation called ``transport hell''.

However, even at the GAT level of abstraction, there are some choices
to be made about the instantiation operation. It is usually convenient
to use parallel substitutions, which means that the instantiation
operation replaces all variables in a context by terms in another
context \emph{at the same time}, that is, $t[x_0\mapsto
  u_0,\dots,x_n\mapsto u_n]$ where $x_0,\dots,x_n$ are all the free
variables in $t$. $(x_0\mapsto u_0,\dots,x_n\mapsto u_n)$ is called a
parallel substitution, and the terms $u_0,\dots,u_n$ are all in the
same context. Parallel substitutions are very natural in the algebraic
setting because they form the morphisms in a category where objects
are contexts (lists of types), and then it is clear what equations
they need to satisfy. Some examples of algebraic definitions of type
theory using parallel substitutions ranging from the more syntactic to
the more semantic: Ehrhard's notion \cite{ehrhard,coquandEhrhard},
categories with families (CwF
\cite{DBLP:conf/types/Dybjer95,Castellan2021}), Awodey's natural
models \cite{DBLP:journals/mscs/Awodey18}, contextual categories
\cite{DBLP:journals/apal/Cartmell86}, C-systems
\cite{DBLP:journals/lmcs/AhrensLV18}, locally cartesian closed
categories \cite{DBLP:journals/mscs/ClairambaultD14}, path categories
\cite{10.1145/3204492}.

In traditional extrinsic presentations
(e.g.\ \cite{alti:phd93,DBLP:books/daglib/0005958}) instantiation
simply replaces all free occurrences of a variable $x$ with a term $u$
as in the notation $t[x\mapsto u]$. A substitutions consists of a
variable name (a pointer into a context) and a term. Substitutions
cannot be composed, thus they do not form a category. To bridge the
gap between the intrinsic parallel substitution world and the
extrinsic single substitution world, Voevodsky introduced B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} which are an algebraic
desription of a single substitution calculus. B-systems involve
complex rules describing telescopes, weakenings and substitutions
under telescopes, with several equations.

In this paper, we introduce a new single substitution calculus (SSC)
which is simpler and more minimalistic than the existing algebraic
approaches. The instantiation operation $\blank[\blank]$ takes either a
single substitution or a single weakening as an argument, and there
are eight equations explaining how these behave: four of these say how
to instantiate variables and the other four equations are needed to
typecheck the first four. The SSC shows that to explain type theory in
an algebraic way, there is no need for categories, parallel
substitutions or weakenings, empty substitution, telescopes, or
combinations of these. All our operations are easy to motivate, we
showcase this in Section \ref{sec:tt} which is a tutorial introduction
to our theory. We believe that our SSC is in a sweet spot: our
notations are close to traditional extrinsic notations, but they are
algebraic, thus come with a well behaved model theory, and can be
easily related to more semantic descriptions.

Our SSC is in some sense too minimalistic: for example, the equation
$b[\p^+][\langle\q\rangle] = b$ is not \emph{derivable} in any model,
but it is \emph{admissible} in the syntax. (Here $b : \Tm\,(\Gamma\ext
A)\,B$ is a term which depends on a context $\Gamma$ and an extra
variable of type $A$; $b[\p^+] : \Tm\,(\Gamma\ext A\ext
A[\p])\,(B[\p^+])$ is a version where we weakened \emph{before} the
last variable; in $b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$
we substituted the last variable for the previous one.)  This is
analogous to e.g.\ parametricity results
\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
arbitrary model, but they hold in the syntax. In this paper, we show
that the \emph{syntax} of our single substitution theory and the
syntax of the corresponding parallel substitution theory (CwF-based
theory) are isomorphic. All CwF-based models are models of our single
substitution theory, but not necessarily the other way. While
B-systems are equivalent to CwFs (and C-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more SSC models
than CwF-based models. This relationship is similar to that of
extensional combinatory calculus and lambda calculus, where the
syntaxes are equivalent, and every lambda model is a combinatory
model, but not the other way round
\cite{DBLP:conf/fscd/AltenkirchKSV23}. Another example is the
relationship between monoids over a set $X$ and nil/cons algebras over
$X$, where every monoid is a nil/cons-algebra, but not the other way;
but it is well-known that free monoids over $X$ (syntax of monoid over
$X$) are the same as lists of $X$ (the syntax of nil/cons algebras
over $X$).

We introduce a technology for proving admissible rules in our
theory. We define $\alpha$-normal forms and show
$\alpha$-normalisation: every term admits an $\alpha$-normal form.
Thus, when one wants to prove something about all terms, it is enough
to prove it for $\alpha$-normal forms. Just as $\beta$-normal forms
don't distinguish between $\beta$-convertible terms, $\alpha$-normal
forms don't include explicit substitutions. $\alpha$-normal forms are
not like ordinary normal forms because they are still quotiented by
the computation/uniqueness rules like $\beta$/$\eta$ for function
space. We use induction on $\alpha$-normal forms to define parallel
substitutions and prove all the rules of CwFs in the syntax of SSC.

Our SSC has 8 equations, and these are enough to describe type
theories with arbitrary choice of type formers (see the description of
SOGATs in Subsection \ref{sec:related}). In the particular case of
type theory with a Coquand-universe and $\Pi$ types, we can do better:
4 (conditional) equations on variables are enough, because using the
fact that every type is represented by a terma, we can derive the other
4 equations.

If in addition to having a universe and $\Pi$ types, our type theory
also supports unit and $\Sigma$ types, we can actually derive a
CwF-based model from an SSC-model. The idea is that $\Sigma$ types can
represent contexts, functions between $\Sigma$ types represent
parallel substitutions, functions into the universe represent depenent
types. Then the SSC is just there to formulate $\Sigma$, $\Pi$, $\U$,
and then we use these to define a parallel substitution calculus. So
although SSC models and CwFs without any type formers are not
equivalent, if we equip them with enough type formers, they become
(weakly) equivalent.

All of our constructions can be understood as happening internal to
extenional type theory, and most results in this paper are formalised
in Agda. Most results in this paper were presented at the TYPES 2024
conference \cite{singleTypes}.

\begin{itemize}
\item A new generalised algebraic presentation of type theory in the
  form of a minimalistic single substitution calculus. We show that
  there is no need for parallel substitutions, empty substitution,
  parallel weakenings, telescopes, or combinations of these to
  describe the syntax of type theory in an intrinsic way. We present
  our calculus in an easy-to-understand way.
\item The $\alpha$-normalisation technique and deriving parallel
  substitutions from the single substitution syntax.
\item For type theory with $\Pi$, $\U$, a minimised presentation of
  the equations which results in an isomorphic theory.
\item For type theory with $\Pi$, $\U$, $\top$, $\Sigma$, we show that
  single and parallel models are equivalent in a weak sense.
\end{itemize}

@@

After discussing related work, in Section \ref{sec:tt}, we introduce
our single substitution calculus (SSC) only relying on a working
knowledge of an implementation of type theory such as Agda, no prior
experience with the metatheory of type theory is required. Section
\ref{sec:admissible} shows that in the syntax of SSC, several new
equations are admissible, and that the syntax of SSC is actually
isomorphic to the CwF-based syntax. In Section \ref{sec:minimisation}
we minimise our SSC obtaining a theory with fewer equations, relying
on the presence of $\Pi$ and $\U$ in our theory. In Section
\ref{sec:cwf}, we show that in the presence of $\Pi$, $\U$ and
$\Sigma$, SSC-models are actually equivalent to CwF-models in a weak
sense. We conclude in Section \ref{sec:conclusion}.

@@

\paragraph{B-systems} B-systems were introduced by Voevodsky \cite{voevodsky2014bsystems} as

@@

substitutions. B-systems correspond to our SSC as essentially
algebraic theories correspond to generalised algebraic theories, or
sets with a map into $I$ correspond to indexed families over $I$
\cite[page 221]{DBLP:journals/apal/Cartmell86}. Another difference is
that we have fewer and less general equations, resulting in the fact
that we have more models than B-systems. However in the syntax of our
theory, all the rules of B-systems are admissible. We describe the

@@

\paragraph{Ehrhard's calculus}
The first generalised algebraic presentation of type theory was Thomas
Ehrhard's calculus \cite{ehrhard,coquandEhrhard} which featured a
parallel substitution calculus with $\p$, $\blank^+$ and
$\langle\blank\rangle$ operations, just like in our calculus. Our
single substitution calculus is essentially Ehrhard's calculus with
the categorical composition and identity operations
removed. Categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}) feature $\p$, $\q$,
$\blank,\blank$ operations which make it more apparent that
substitutions are lists of terms. CwFs are equivalent to Ehrhard's
calculus, and so are contextual categories/C-systems
\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
the natural models of Awodey \cite{DBLP:journals/mscs/Awodey18} and
B-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.

\paragraph{Formalisations of type theory}
Most computer formalisations of type theory are extrinsic
\cite{DBLP:journals/pacmpl/0001OV18,DBLP:conf/cpp/AdjedjLMPP24,DBLP:journals/jar/SozeauABCFKMTW20}:
they define the syntax as abstract syntax trees, and equip it with
typing and conversion relations. A higher level representation is
working in setoid hell \cite{chapman09eatitself}: terms are
intrinsically well-typed, but conversion is still an explicit
relation. For formalising the GAT-level syntax, one needs a stronger
metatheory than ordinary Agda or Coq: the metatheory has to support
quotient inductive-inductive types (QIITs
\cite{DBLP:journals/pacmpl/KaposiKA19}), in other words, initial
models of GATs. Altenkirch and Kaposi
\cite{DBLP:conf/popl/AltenkirchK16} formalised the syntax of type
theory using postulated QIITs in Agda, together with parametricity and
normalisation \cite{lmcs:4005} proofs. Brunerie and de Boer
\cite{initiality-agda} constructed the initial contextual category in
Agda using postulated quotients. Although there is no general proof
published about this, we have experimental evidence that QIITs are
supported by the cubical set model \cite{DBLP:conf/lics/CoquandHM18}
and the setoid model \cite{kaposi-qiit-setoid}, thus Cubical Agda
\cite{DBLP:journals/jfp/VezzosiMA21} and setoid/observational type
theory \cite{setoid,DBLP:phd/hal/Pujet22} support QIITs. For example,
in Cubical Agda, set-truncated and groupoid-truncated syntaxes of type
theory have been formalised \cite{cohtt}. Without QIITs, tricks such
as shallow embedding the syntax can be used to formalise metatheoretic
results about type theory such as gluing
\cite{kaposi_et_al:LIPIcs:2019:10532}, and its special cases
canonicity and parametricity \cite{kaposi-shallow}.

\paragraph{SOGATs}
We can define languages with binders as second-order theories
following higher-order abstract syntax \cite{DBLP:conf/lics/Hofmann99}
and logical frameworks
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}. Second-order
generalised algebraic theories (SOGATs
\cite{uemura,DBLP:conf/fscd/KaposiX24}) can be used to present type
theory even more abstractly than GATs: they even abstract over the way
of defining the substitution calculus: contexts, substitutions,
instantiations are simply modelled by the metatheoretic function
space. For example, a type theory with $\Pi$, Coquand-universes and a
universe level lifting operation is described by the following SOGAT
(where $\cong$ means isomorphism):
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
\end{alignedat}
\end{equation}
% Adding $\top$ and $\Sigma$ types (Definition \ref{def:sigma}) is
% described as the following extension of the above SOGAT:
% \begin{equation}\label{eq:sigma}
%   \begin{alignedat}{10}
%     & \top && : \Ty\,0 \\
%     & \tt && : \Tm\,\top \\
%     & \top\eta && : t = \tt \\
%     & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
%     & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
% \end{alignedat}
% \end{equation}
Second-order models of SOGATs are not well-behaved as there is no good
notion of homomorphism between them. This is why Kaposi and Xie
\cite{DBLP:conf/fscd/KaposiX24} translate SOGATs to GATs: each GAT
comes with a category of models with initial model (syntax), (co)free
models, and so on. Actually, Kaposi and Xie define two different
translations: one based on parallel substitutions and one based on
single substitutions. If we apply their parallel translation to the
SOGAT
\begin{equation}\label{eq:tytm}
\begin{alignedat}{10}
  & \Ty && : \Set \\
  & \Tm && : \Ty \ra\Set
\end{alignedat}
\end{equation}
with only two sorts and no operations or equations, we obtain the GAT
of category with families (CwF). If we use their single substitution
translation on the SOGAT (\ref{eq:tt}), we obtain Definition
\ref{def:tt} (also listed in Appendix \ref{app:tt}). This paper can be
seen as taking a big enough type theory (described as a SOGAT), and
investigating how the result of its single substitution translation
relates to the result of its parallel translation (the latter
first-order theory is quite well-known).
